**Bayesian statistics for longitudinal studies in biomedical research**

### 2020-11-04 Manuscripts/Manuscript_AM_v1.Rmd
Updated abstract, section 1

Generalized additive models (GAMs) are a subset of generalized linear models that use _smooth functions_ (hencefort _smooths_) to estimate the parameters of a model. They have been used in palaeolimnology, ecology and clinical studies to model longitudinal data[@woolway2016;@hefley2017;@ko2007]. Briefly, GAMs use a combination of multiple functions (basis functions)to construct the smooths of the model [@wood2017]. Their main advantage over LMEMs and rm-ANOVA is that the model specification is directed by the _smooths_ rather than by a parametric relationship. This allows a consistent fit of the model with the data, and estimations of significance using the terms of the model. However, certain assumptions about the data are necessary: a normal distribution and constant variance of the residuals with the mean response. Therefore, GAMs provide a more suitable statistical method to analyze biomedical longitudinal data, when these assumptions are met.

If the assumptions of GAMs do not hold,  the field of _Bayesian statistics_ represents a relatively new area of Statistics that does not rely on _p-values_ and hypothesis tests to analyze information. Bayesian statistics can work with missing observations, allow the data (and not an underlying assumed distribution) to determine the outcome in regard to significance and are able to expand the number of comparisons and inferences derived form the analysis. On the other hand, the shift that Bayesian statistics represent from the traditional _frequentist_ statistical view in research, the computational tools required for its implementation, and the underlying mathematical theory have limited the use of this approach in the biomedical research community. However,Bayesian theory is intuitive and shares some principles with _frequentist_ statistics, and there is an increasing use and recognition of the advantages of their use across different areas of biomedical research such as clinical trial design and imaging[@biswas2009;@kelter2020;@kwon2020;@zhou2017]. Additionaly,the current development in computational tools, specifically the programming language $\textsf{R}$, enable a rapid implementation of Bayesian models.

To this end, the goals of this study are: a) to present the limitations of (rm-ANOVA) and LMEMs over longitudinal data, and demonstrate how these limitations in turn affect the results of the analysis b) introduce in a practical and amenable manner the theory of Bayesian statistics highlighting its applicability to biomedical research and c)Implement a Bayesian analysis over a set of simulated data that matches previously reported trends in longitudinal biomedical studies. With an emphasis on reproducibility by providing the code and dataset used, this will provide biomedical researchers a clear view of the advantages of Bayesian statistics for the analysis of longitudinal data.

@@

#### 1.1 the _frequentist_ case for constant correlation (ANOVA limitations)

The first 
Missing observations, and correlation between measurements. How rm-ANOVA is limited by missing observations, and how both rm-ANOVA and LMEMs are limited with data that does not follow a linear trend (equations for both situations and the fit they produce.)

>>>>

Generalized additive models (GAMs) are a subset of generalized linear models that use _smooth functions_ (hencefort _smooths_) to estimate the parameters of a model. They have been used in palaeolimnology, ecology and clinical studies to model longitudinal data[@woolway2016;@hefley2017;@ko2007]. Briefly, GAMs use a combination of multiple functions (basis functions)to construct the smooths of the model [@wood2017]. Their main advantage over LMEMs and rm-ANOVA is that the model specification is directed by the _smooths_ rather than by a parametric relationship. This allows a consistent fit of the model with the data, and estimations of significance using the terms of the model. However, certain assumptions about the data are necessary: a normal distribution and constant variance of the residuals with the mean response. Therefore, GAMs provide a more suitable statistical method to analyze biomedical longitudinal data, when these assumptions of the model are met by the data.

However, it is possible that the assumptions of GAMs for longitudinal data do not hold under  certain circumstances. In that case, the field of _Bayesian statistics_ represents a relatively new area of Statistics that does not rely on _p-values_ and hypothesis tests to analyze information. Bayesian statistics can work with missing observations, allow the data (and not an underlying assumed distribution) to determine the outcome in regard to significance and are able to expand the number of comparisons and inferences derived form the analysis. On the other hand, the shift that Bayesian statistics represent from the traditional _frequentist_ statistical view in research, the computational tools required for its implementation, and the underlying mathematical theory have limited the use of this approach in the biomedical research community. However,Bayesian theory is intuitive and shares some principles with _frequentist_ statistics, and there is an increasing use and recognition of the advantages of their use across different areas of biomedical research such as clinical trial design and imaging[@biswas2009;@kelter2020;@kwon2020;@zhou2017]. Additionally,the current development in computational tools, specifically the programming language $\textsf{R}$, enable a rapid implementation of Bayesian models for longitudinal data.

Therefore, this study focuses in three areas in the analysis of longitudinal data from a biomedical perspective. First, it presents the limitations of (rm-ANOVA) and LMEMs over longitudinal data, and explains how these limitations in turn affect the results of the analysis. Secondly, it uses simulation to generate non-linear longitudinal data and presents the implementation of GAMs as a statistical tool that enables the analysis of data that shows this behavior. And finally, it introduces Bayesian statistics and presents their implementation with GAMs over simulated data to demonstrate the differences and benefits of this approach . With an emphasis on reproducibility by providing the code and dataset used, this study provides biomedical researchers with a clear view of the advantages of different approaches in the statistical analysis of longitudinal data.

@@

### 1 The _frequentist_ case for longitudinal data

While the _repeated measures analysis of variance_ (rm-ANOVA) is the standard for the statistical analysis of longitudinal data, there are key assumptions that are made in order to make the model valid. From a practical view, they can be divided in three areas: linear relationship between covariates and response, constant correlation between measurements, and complete observations for all subjects. Each one of these assumptions is discussed below.

#### 1.1 Linear relationship

When a continuous and normally distributed response $y$ is measured from subject $n$, the basic model for rm-ANOVA becomes[@davis2002]:

$$ y_{ij}=\mu_{ij}+\pi_{ij}+e_{ij}$$
Missing observations, and correlation between measurements. How rm-ANOVA is limited by missing observations, and how both rm-ANOVA and LMEMs are limited with data that does not follow a linear trend (equations for both situations and the fit they produce.)
#### 1.1 Constant correlation

While the _repeated measures analysis of variance_ (rm-ANOVA) is the norm in the biomedical community to analyze longitudinal data, two frequent situations that arise in longitudinal studies limit its applicability. 
Missing observations, and correlation between measurements. How rm-ANOVA is limited by missing observations, and how both rm-ANOVA and LMEMs are limited with data that does not follow a linear trend (equations for both situations and the fit they produce.)
